Using generalized additive models to analyze biomedical non-linear longitudinal data: _Beyond Linear Mixed Models and Repeated Measures ANOVA_

### 2021-07-09 Manuscripts/Manuscript by chapters - Wiley submission/00-Full_document_SIM_No_Appendix.tex
Created folder for Wiley submission

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Using generalized additive models to analyze biomedical non-linear longitudinal data},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{listings}
\newcommand{\passthrough}[1]{#1}
\lstset{defaultdialect=[5.3]Lua}
\lstset{defaultdialect=[x86masm]Assembler}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\usepackage{helvet}
\usepackage{lineno}
\usepackage[affil-it,blocks]{authblk}
\usepackage{hyperref}
\usepackage{graphicx}
\usepackage{setspace}
\usepackage[nomarkers,figuresonly]{endfloat}
%set a box to put the ORCID logo
\newbox{\myorcidaffilbox}
\sbox{\myorcidaffilbox}{\large\includegraphics[height=1.7ex]{orcid}}

@@

\title{\textbf{Using generalized additive models to analyze biomedical non-linear longitudinal data}}
\usepackage{etoolbox}
\makeatletter
\providecommand{\subtitle}[1]{% add subtitle to \maketitle
  \apptocmd{\@title}{\par {\large #1 \par}}{}{}
}
\makeatother
\subtitle{\emph{Beyond repeated measures ANOVA and Linear Mixed Models}}
\date{\vspace{-2.5em}}

@@

In contrast, \emph{generalized additive models} (GAMs) relax the linearity assumption, and allow the data to determine the fit of the model while permitting missing observations and different correlation structures. Therefore, GAMs present an excellent choice to analyze non-linear longitudinal data in the context of biomedical research. This paper summarizes the limitations of rm-ANOVA and LMEMs and uses simulated data to visually show how both methods produce biased estimates when used on non-linear data. We also present the basic theory of GAMs, and using trends of oxygen saturation in tumors reported in the biomedical literature, we simulate example longitudinal data (2 treatment groups, 10 subjects per group, 5 repeated measures for each group) to demonstrate how these models can be computationally implemented. We show that GAMs are able to produce estimates that are consistent with the trends of biomedical non-linear data even in the case when missing observations exist (with 40\% of the simulated observations missing), allowing reliable inference from the data. To make this work reproducible, the code and data used in this paper are available at: \url{https://github.com/aimundo/GAMs-biomedical-research}.

@@

{\centering \includegraphics[width=0.75\linewidth]{00-Full_document_No_Appendix_files/figure-latex/l-q-response-1}

@@

{\centering \includegraphics[width=0.75\linewidth]{00-Full_document_No_Appendix_files/figure-latex/basis-plot-1}

@@

{\centering \includegraphics[width=0.75\linewidth]{00-Full_document_No_Appendix_files/figure-latex/sim-smooth-plot-1}

@@

{\centering \includegraphics[width=0.75\linewidth]{00-Full_document_No_Appendix_files/figure-latex/plot-pairwise-comp-1}

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Generalized additive models to analyze biomedical non-linear longitudinal data in R: Beyond repeated measures ANOVA and Linear Mixed Models},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{listings}
\newcommand{\passthrough}[1]{#1}
\lstset{defaultdialect=[5.3]Lua}
\lstset{defaultdialect=[x86masm]Assembler}
\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\usepackage{mathptmx}
\usepackage{lineno}
\usepackage[affil-it,blocks]{authblk}
\usepackage{hyperref}
\usepackage{graphicx}
\usepackage{setspace}
%\usepackage[nomarkers,figuresonly]{endfloat}
%set a box to put the ORCID logo
\newbox{\myorcidaffilbox}
\sbox{\myorcidaffilbox}{\large\includegraphics[height=1.7ex]{orcid}}

@@

\title{\textbf{Generalized additive models to analyze biomedical non-linear longitudinal data in R: Beyond repeated measures ANOVA and Linear Mixed Models}}
\date{\vspace{-2.5em}}

@@

In contrast, generalized additive models (GAMs) relax the linearity assumption, and allow the data to determine the fit of the model while permitting missing observations and different correlation structures. Therefore, GAMs present an excellent choice to analyze non-linear longitudinal data in the context of biomedical research. This paper summarizes the limitations of rm-ANOVA and LMEMs and uses simulated data to visually show how both methods produce biased estimates when used on non-linear data. We present the basic theory of GAMs, and using reported trends of oxygen saturation in tumors we simulate example longitudinal data (2 treatment groups, 10 subjects per group, 5 repeated measures for each group) to demonstrate their implementation in R. We also show that GAMs are able to produce estimates that are consistent with the trends of non-linear data even in the case when missing observations exist (with 40\% of the simulated observations missing). To make this work reproducible, the code and data used in this paper are available at: \url{https://github.com/aimundo/GAMs-biomedical-research}.

@@

{\centering \includegraphics[width=0.75\linewidth]{00-Full_document_SIM_No_Appendix_files/figure-latex/l-q-response-1}

@@

{\centering \includegraphics[width=0.75\linewidth]{00-Full_document_SIM_No_Appendix_files/figure-latex/basis-plot-1}

@@

{\centering \includegraphics[width=0.75\linewidth]{00-Full_document_SIM_No_Appendix_files/figure-latex/sim-smooth-plot-1}

@@

{\centering \includegraphics[width=0.75\linewidth]{00-Full_document_SIM_No_Appendix_files/figure-latex/plot-pairwise-comp-1}

### 2021-07-09 Manuscripts/Manuscript by chapters - Wiley submission/00-OUP_Appendix.Rmd
Created folder for Wiley submission

---
title: '**Using generalized additive models to analyze biomedical non-linear longitudinal data**'
subtitle: _Beyond repeated measures ANOVA and Linear Mixed Models_
header-includes:
    \usepackage{placeins}
output:  
  bookdown::pdf_document2:
    #template: my-template.tex #if a custom template that removes the additional "and" in the author information is desired
    pandoc_args: --listings #calls the listings package to fit code within the page margins
    keep_tex: yes #keep LaTeX file for submission
    fig_caption: yes #allows captions in figures
    toc: false #do not include table of contents
    extra_dependencies:
      subfig: null #allows for subfigures
      breqn: null #line breaks for long equations
      caption: ["font={small}"] #size of the figure captions
      #float: null #allows for control of placement of figures
    includes:
      in_header: preamble_OUP.sty #additional LaTeX formatting
  bookdown::word_document2:
    fig_caption: yes #figure caption
    keep_md: yes
  bookdown::html_document2:
    css: "style.css" #style for the HTML document
csl: elsevier-with-titles.csl #style for references
bibliography: refs.bib #references
link-citations: yes #adds links to the citations
'': default
---

# Abstract

@@

<!-- ```{r child='07-Appendix.Rmd'} -->

<!-- ``` -->

>>>>

---
title: |
    | **Using generalized additive models to analyze biomedical non-linear longitudinal data**
    | _Beyond repeated measures ANOVA and Linear Mixed Models_
    | 
    | SUPPLEMENTARY MATERIALS: APPENDIX
header-includes:
    \usepackage{placeins}
output:  
  bookdown::pdf_document2:
    #template: my-template.tex #if a custom template that removes the additional "and" in the author information is desired
    pandoc_args: --listings #calls the listings package to fit code within the page margins
    keep_tex: yes #keep LaTeX file for submission
    fig_caption: yes #allows captions in figures
    toc: false #do not include table of contents
    extra_dependencies:
      subfig: null #allows for subfigures
      breqn: null #line breaks for long equations
      caption: ["font={small}"] #size of the figure captions
      #float: null #allows for control of placement of figures
    includes:
      in_header: preamble_OUP_Appendix.sty #additional LaTeX formatting
  bookdown::word_document2:
    fig_caption: yes #figure caption
    keep_md: yes
  bookdown::html_document2:
    css: "style.css" #style for the HTML document
csl: elsevier-with-titles.csl #style for references
bibliography: refs.bib #references
link-citations: yes #adds links to the citations
'': default
---

\newpage
# Abstract

@@

```{r child='07-Appendix.Rmd'}

```

### 2021-07-09 Manuscripts/Manuscript by chapters - Wiley submission/Full_document_SIM_rticles_No_Appendix.Rmd
Created folder for Wiley submission

---
title: '**Using generalized additive models to analyze biomedical non-linear longitudinal data**'
subtitle: _Beyond repeated measures ANOVA and Linear Mixed Models_
header-includes:
    \usepackage{placeins}
output:  
  bookdown::pdf_document2:
    #template: my-template.tex #if a custom template that removes the additional "and" in the author information is desired
    pandoc_args: --listings #calls the listings package to fit code within the page margins
    keep_tex: yes #keep LaTeX file for submission
    fig_caption: yes #allows captions in figures
    toc: false #do not include table of contents
    extra_dependencies:
      subfig: null #allows for subfigures
      breqn: null #line breaks for long equations
      caption: ["font={small}"] #size of the figure captions
      #float: null #allows for control of placement of figures
    includes:
      in_header: preamble_OUP.sty #additional LaTeX formatting
  bookdown::word_document2:
    fig_caption: yes #figure caption
    keep_md: yes
  bookdown::html_document2:
    css: "style.css" #style for the HTML document
csl: biostatistics.csl #style for references
bibliography: refs.bib #references
link-citations: yes #adds links to the citations
'': default
---

# Summary

In biomedical research, the outcome of longitudinal studies has been traditionally analyzed using the _repeated measures analysis of variance_ (rm-ANOVA) or more recently, _linear mixed models_ (LMEMs). Although LMEMs are less restrictive than rm-ANOVA in terms of correlation and missing observations, both methodologies share an assumption of linearity in the measured response, which results in biased estimates and unreliable inference when they are used to analyze data where the trends are non-linear, which is a common occurrence in biomedical research.

In contrast, _generalized additive models_ (GAMs) relax the linearity assumption, and allow the data to determine the fit of the model while permitting missing observations and different correlation structures. Therefore, GAMs present an excellent choice to analyze non-linear longitudinal data in the context of biomedical research. This paper summarizes the limitations of rm-ANOVA and LMEMs and uses simulated data to visually show how both methods produce biased estimates when used on non-linear data. We also present the basic theory of GAMs, and using trends of oxygen saturation in tumors reported in the biomedical literature, we simulate example longitudinal data (2 treatment groups, 10 subjects per group, 5 repeated measures for each group) to demonstrate how these models can be computationally implemented. We show that GAMs are able to produce estimates that are consistent with the trends of biomedical non-linear data even in the case when missing observations exist (with 40% of the simulated observations missing), allowing reliable inference from the data. To make this work reproducible, the code and data used in this paper are available at: https://github.com/aimundo/GAMs-biomedical-research.

# Keywords {-}

cancer biology; tumor response; generalized additive models; simulation; R

>>>>

---
title: "Using generalized additive models to analyze biomedical non-linear longitudinal data: _Beyond Linear Mixed Models and Repeated Measures ANOVA_"
author:
- name: Ariel I. Mundo
  num: a
- name: Timothy J. Muldoon*
  num: a
- name: John R. Tipton
  num: b
address:
- num: a
  org: Department of Biomedical Engineering, University of Arkansas, Fayetteville, AR, US
- num: b
  org: Department of Mathematical Sciences, University of Arkansas, Fayetteville, AR, US
corres: "*Timothy J. Muldoon, Engineering Research Center Suite 2540, Fayetteville, AR, 72701 \\email{tmuldoon@uark.edu}"
presentaddress: This is sample for present address text this is sample for present address text
authormark: Mundo \emph{et al}.
articletype: Tutorial in Biostatistics
received: 
revised: 
accepted: 
abstract: "In biomedical research, the outcome of longitudinal studies has been traditionally analyzed using the _repeated measures analysis of variance_ (rm-ANOVA) or more recently, _linear mixed models_ (LMEMs). Although LMEMs are less restrictive than rm-ANOVA in terms of correlation and missing observations, both methodologies share an assumption of linearity in the measured response, which results in biased estimates and unreliable inference when they are used to analyze data where the trends are non-linear, which is a common occurrence in biomedical research. In contrast, _generalized additive models_ (GAMs) relax the linearity assumption, and allow the data to determine the fit of the model while permitting missing observations and different correlation structures. Therefore, GAMs present an excellent choice to analyze non-linear longitudinal data in the context of biomedical research. This paper summarizes the limitations of rm-ANOVA and LMEMs and uses simulated data to visually show how both methods produce biased estimates when used on non-linear data. We also present the basic theory of GAMs, and using trends of oxygen saturation in tumors reported in the biomedical literature, we simulate example longitudinal data (2 treatment groups, 10 subjects per group, 5 repeated measures for each group) to demonstrate how these models can be computationally implemented. We show that GAMs are able to produce estimates that are consistent with the trends of biomedical non-linear data even in the case when missing observations exist (with 40% of the simulated observations missing), allowing reliable inference from the data. To make this work reproducible, the code and data used in this paper are available at: https://github.com/aimundo/GAMs-biomedical-research."
keywords: Longitudinal data; Biomedical data; Simulation; R;
bibliography: refs.bib
output:
  bookdown::pdf_book:
    pandoc_args: [ "--listings" ]
    base_format: rticles::sim_article
    keep_tex: yes
    fig_caption: yes
    includes:
      in_header: preamble.sty #LaTeX formatting options
---
